---
title: "NOAA vignette Mastering Software Development in R Capstone"
author: "Adrian Accurso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NOAAvignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

After installing the package with install.packages("NOAA"), load the library with the following command:

```{r setup}
library(NOAA)
```

```{r, echo = FALSE, include = FALSE}
library(NOAA)
library(dplyr)
library(ggplot2)
library(readr)
library(grid)
library(lubridate)
library(leaflet)
library(scales)
```

This vignette gives a brief overview of the `NOAA` R package created for the purpose of visualizing NOAA earthquake data. It processes data from [NOAA database](https://www.ngdc.noaa.gov/hazel/view/hazards/earthquake/search) and visualizes them using `ggplot2` and `leaflet` packages. 

## Installation

To install the package type the following (to work with the development version from Github):

```
library(devtools)
install_github("adrian-accurso/NOAA")
library(NOAA)
```

## Package functions

6 functions are exported and available for the users:

- `eq_clean_data()`
- `geom_timeline()`
- `geom_timeline_label()`
- `theme_timeline()`
- `eq_create_label()`
- `eq_map()`

Please find a short description with examples on how to use the functions. The example data from NOAA can be found in the package directory in the `\importeddata` folder.

## Clean data

The first function is used to clean data for the visualization and so that the category names are consistent with later used functions. It creates a DATE column in `Date` format, transforms latitude and longitude to numeric format and trims country from LOCATION_NAME.

```{r eq_read_example, message = FALSE}
file_name <- system.file("vignettes/earthquakesFile2.txt", package = "NOAA")
data <- readr::read_delim(file_name, delim = "\t")
eq_clean_data(data)
```

## Visualize earthquake timeline

The next three functions utilize the `ggplot2` package to visualize earthquake timeline. The basic `geom_timeline()` geom requires clean data instead of raw data as downloaded from the NOAA website. The required aesthetics is `x` with dates, optional are `y` for grouping by country, and `size` and `color` that can be applied according to user needs. The `geom_timeline_label()` function requires additional `label` aesthetic for labeling. For better visualization of these two geoms, `theme_timeline()` theme was added. An example follows:

```{r eq_timeline_example, fig.width = 7, fig.height = 4}
data %>% eq_clean_data() %>%
     filter(COUNTRY %in% c("MEXICO", "GUATEMALA"), YEAR > 1900) %>%
     ggplot(aes(x = DATE,
                y = COUNTRY,
                color = as.numeric(TOTAL_DEATHS),
                size = as.numeric(EQ_PRIMARY)
     )) +
     geom_timeline() +
     geom_timeline_label(aes(label = LOCATION_NAME), n_max = 5) +
     theme_timeline() +
     labs(size = "Richter scale value", color = "# deaths")
```

## Visualize earthquakes on map

This package employs `leaflet` functions in the `eq_map()` function to produce maps of the country of interest and its earthquakes. The map is cropped based on the coordinates of the supplied country. The geom can use an optional `eq_create_label()` function to provide more information about the dates and severity of earthquakes if the user desires. The output is an interactive map with detailed clickable points for each earthquake:

```{r eq_map_example, fig.width = 7, fig.height = 4}
data %>% 
  eq_clean_data() %>% 
  dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>% 
  dplyr::mutate(popup_text = eq_create_label(.)) %>% 
  eq_map(annot_col = "popup_text")
```

